---
title: "Introduction to adestr"
author: "Jan Meis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to adestr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This package implements methods to evaluate the performance characteristics
of various point and interval estimators for adaptive two-stage designs with 
prespecified sample-size recalculation rules. Further, it allows for
evaluation of these estimators on real datasets, and it implements methods
to calculate p-values.

Currently, it works for designs objects which were produced by the
R-package `adoptr`, which calculates optimal design parameters for
adaptive two-stage designs. You can learn about adoptr here:
[optad.github.io/adoptr/](https://optad.github.io/adoptr/).

An introductory vignette covering common usecases is given at [https://jan-imbi.github.io/adestr/articles/Introduction.html](https://jan-imbi.github.io/adestr/articles/Introduction.html).

This package comes suite of unit tests. The code of the test cases can be viewed here: [https://github.com/jan-imbi/adestr/tree/master/tests/testthat](https://github.com/jan-imbi/adestr/tree/master/tests/testthat). The authors assume no responsibility for the correctness of the
code or results produced by its usage. Use at your own risk.

# Fitting a design with adoptr
In order to showcase the capabilities of this package, we need a trial design first.
We refer to [the example from the adoptr documentation](https://optad.github.io/adoptr/articles/adoptr.html)
for this.
You can read more about optimal adaptive designs fitted via the adoptr package here:
[optad.github.io/adoptr/articles/adoptr_jss.html](https://optad.github.io/adoptr/articles/adoptr_jss.htmll).

For the sake of this introduction, a pre-computed version of the first example from
[optad.github.io/adoptr/articles/adoptr.html](https://optad.github.io/adoptr/articles/adoptr.html)
is provided with this package via the `get_example_design` function.

```{r}
library(adestr)
get_example_design(two_armed = TRUE)
```

# Example
## Evaluating the mean squared of an estimator
Now that we have created an optimal adaptive design, we can investigate the performance
characteristics of various estimators for the mean in that design.
To this end, the `evaluate_estimator` function can be used.

```{r}
evaluate_estimator(
 score = MSE(),
 estimator = SampleMean(),
 data_distribution = Normal(two_armed = TRUE),
 design = get_example_design(two_armed = TRUE),
 mu = 0.3,
 sigma = 1
)
```

The mean squared error of the sample mean depends on the true underlying value of the
paramter $\mu$, which of course is unknown. Therefore, it may be
advisable to use the `evaluate_estimator` function on an array of values for $\mu$
to investigate the distributional properties of an estimator.

In the following, the MSE of the sample mean vs. a weighted sample mean with fixed weights
will be plotted.

```{r, fig.width=7.2, fig.height=4, dev="svg"}
mse_mle <- evaluate_estimator(
  score = MSE(),
  estimator = SampleMean(),
  data_distribution = Normal(two_armed = TRUE),
  design = get_example_design(two_armed = TRUE),
  mu = seq(-0.75, 1.32, .03),
  sigma = 1
)
mse_weighted_sample_means <- evaluate_estimator(
  score = MSE(),
  estimator = WeightedSampleMean(w1 = .8),
  data_distribution = Normal(two_armed = TRUE),
  design = get_example_design(two_armed = TRUE),
  mu = seq(-0.75, 1.32, .03),
  sigma = 1
)
plot(c(mse_mle, mse_weighted_sample_means))
```

## Analyzing datasets
Next, let us look at how to the package can be used to calculate estimates
after data has been collected.

The first stage data of a trial might look like this:

```{r}
set.seed(321)
dat <- data.frame(
 endpoint = c(rnorm(56, .3, 1), rnorm(56, 0, 1)),
 group = factor(rep(c("trt", "ctl"),
                    c(56,56)), levels = c("trt", "ctl")),
 stage = rep(1, 112)
)
head(dat)
```

```{r}
analyze(data = dat,
        statistics = list(),
        data_distribution = Normal(two_armed = TRUE),
        design = get_example_design(two_armed = TRUE),
        sigma = 1)
```
The results suggest recruiting 23 more patients per group for the second stage.

We will simulate 47 more patients per group:

```{r}
dat <- rbind(dat,
             data.frame(
               endpoint = c(rnorm(47, .3, 1), rnorm(47, 0, 1)),
               group = factor(rep(c("trt", "ctl"),
                                  c(47, 47)), levels = c("trt", "ctl")),
               stage = rep(2, 94)
             ))
```

Now, we can use the `analyze` function to evaluate a selection of estimators on the complete dataset:

```{r}
analyze(
 data = dat,
 statistics = c(
   SampleMean(),
   BiasReduced(),
   PseudoRaoBlackwell(),
   MedianUnbiasedStagewiseCombinationFunctionOrdering(),
   StagewiseCombinationFunctionOrderingCI(),
   StagewiseCombinationFunctionOrderingPValue()
   ),
 data_distribution = Normal(two_armed = TRUE),
 sigma = 1,
 design = get_example_design(two_armed = TRUE)
)
```

The estimates presented here are for the difference in means of the two normal distributions.
Keep in mind that a difference of $\mu=0.3$ was used in the simulation.

Note that while the median unbiased estimator performs well in this particular example, this
is not universally true.

