---
title: "Introduction to adestr"
author: "Jan Meis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to adestr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# [This documentation is work in progress]

This package implements methods to evaluate the performance characteristics
of various point and interval estimators for adaptive two-stage designs with 
prespecified sample-size recalculation rules. Further, it allows for
evaluation of these estimators on real datasets, and it implements methods
to calculate p-values.

Currently, it works for designs objects which were produced by the
R-package `adoptr`, which calculates optimal design parameters for
adaptive two-stage designs. You can learn about adoptr here:
[kkmann.github.io/adoptr/](https://kkmann.github.io/adoptr/).


## Examples
### Evaluating the mean squared error of an estimator

```{r}
library(adestr)
evaluate_estimator(
 score = MSE(),
 estimator = SampleMean(),
 data_distribution = Normal(TRUE),
 design = get_example_design(),
 mu = c(0.3),
 sigma = 1
)
```


### Analyzing datasets
```{r}
set.seed(321)
dat <- data.frame(
 endpoint = c(rnorm(28, .2, 1), rnorm(28, 0, 1),
              rnorm(23, .2, 1), rnorm(23, 0, 1)),
 group = factor(rep(c("ctl", "trt", "ctl", "trt"),
                    c(28,28,23,23))),
 stage = rep(c(1L, 2L), c(56, 46))
)
analyze(
 data = dat,
 estimator = c(get_example_estimators()),
 data_distribution = Normal(),
 sigma = 1,
 design = get_example_design()
)
```

